Computer Science, Vignettes, and Citation Analysis

### 2019-12-16 Scientometrics/comp.tex
update

\title{Citation Analysis of Computer Science Publications}
%\thanks{Grants or other notes
%about the article that should go on the front page should be
%placed here. General acknowledgments should be placed at the end of the article.}
\subtitle{Relationship To Other Fields}
%\titlerunning{Short form of title}        % if too long for running head

@@

@@

Other classification systems are available such as the All Science Journal Classification (ASJC) developed and maintained by Scopus, the Web of Science research and categorical classifications from Clarivate Analytics, and the NSF classification system~\cite{nsf_classification} [add citations for Scopus and Web of Science]. All three rely on applying one or more journal-derived labels to articles.  A natural consequence is limited specificity at the article level.

Scopus and the Web of Science also include comprehensive bibliographies with citation links and proprietary unique identifiers for publications, some of which can be matched to digital object identifiers (DOIs). However, Shu and colleagues recently noted in a comparative study of the Chinese Science Citation Database (CSCD) and the Web of Science, that 46\% of articles did not belong to the discipline of the journal they were published in~\cite{shu_comparing_2019}. Others have also discussed and critiqued disciplinary assignments using journal based classifications~\cite{wang_large-scale_2016,perianes-rodriguez_comparison_2017}. Article classification systems have been constructed, which escape some of the criticisms of journal-based classification~\cite{traag_louvain_2019,boyack_classification_2014,waltman_new_2012} but seem to be in the research domain rather than the domain of practical use [need to check on this].[May need another bridging sentence to the next paragraph].

@@

\subsection{Data}
 A stable release of the DBLP computer science bibliography~\cite{dblp_ref} consisting of 7,079,994 records was downloaded as dblp-2018-08-01.xml.gz. Just over 95\% of the publications within were published after 1996. Publications were parsed from the XML source file and loaded into a PostgreSQL database.  We have previously parsed the Scopus dataset of over 72,000,000 publications into a custom schema as part of implementing a larger data platform for research evaluation~\cite{GithubERNIE2019}. Records in the DBLP dataset were matched to Scopus identifiers using digital object identifiers (DOIs). This procedure resulted in a dataset of 2,685,356 DBLP publications with Scopus identifiers where 1,278,322 (47.6\%) were labeled as article and 1,407,034 (52.4\%) as conference proceedings. References cited by these publications were then extracted from Scopus (7,129,006 records), resulting in a total of 8,000,411 nodes (publications and references) and 44,296,381 edges that represent citations within the dataset. This dataset is referred to as \emph{comp} (Fig~\ref{fig:ar_cp_annotation}, Table~\ref{tab:comp}).

@@

Figure~\ref{fig:heatmap} permits examination of these data from two perspectives: (i) rows: the clusters that map to a given ASJC minor subject area (ii) columns: ASJC minor subject areas that comprised at least 15\% of the publication in a cluster. Under these conditions of clustering and a threshold of 15\%, 31 of the 334 ASJC minor subject areas are represented. Unsurprisingly, the broad categories Computer Science Applications, Software, and Electrical and Electronic Engineering are seen in 17, 12, and 10 clusters respectively while Artificial Intelligence mapped to 7 different clusters. At the other end of the range, 12 of the 31 ASJC minor subject areas were each detected in a single cluster.

@@

To us, these data suggest that areas central to computer science in 2019 (Salton and Bergmark's historical triad of hardware, software, and theory) are more likely to be found in multiple clusters, whereas areas such as biology, chemistry, and statistics are more restricted in distribution.  A second inference is that, in some cases journal-based classification and our article clusters are similar in terms of article content (Hardware and Architecture). Thirdly, that some areas of computer science are relatively ubiquitous, e.g. computer applications. Lastly, that interactions with fields outside computer science are restricted to a single cluster or small numbers of clusters, e.g., cluster 17  and cluster 18.

We also attempted to map these 20 clusters assembled through direct citation to the CCS classification by manually matching the top 25-50 most heavily cited publications in each cluster to corresponding categories in the CCS. This was feasible in some cases, clusters 0 and 1 mapped reasonably well to the top level categories Hardware and Computer Systems Organization but much less so in the cases of other clusters. Further, the top cited papers often derived from biology and we were concerned that they were not representative of the cluster. The challenges inherent in labeling citation clusters at a high level of aggregation have been previously noted~\cite{waltman_new_2012}.

@@

\subsection{Clustering by Co-citation.} Co-citation was first described independently by Small and Marshakova in 1973~\cite{small_co-citation_1973,marshakova-shaikevich_co-citation_1973}, and provides insight into the emergence of new ideas derived from the association of previously independent ones.  Co-citation thus provides some degree of evolutionary insight and has been used for clustering publications~\cite{boyack_cocitation_2010,boyack_improving_2013,small_structure_1974,small_clustering_1985}. Unlike, clustering by direct citation using Graclus, which preserves all publications in \emph{comp}, the process of clustering by co-citation is progressively selective. As described in Methods, we first identified the top 10\% most-cited articles in the `\emph{comp} dataset and then documented all co-cited pairs of their cited references that were also elements of \emph{comp}. The frequency of co-citation was then computed across Scopus \emph{comp} and normalized using Salton's correction. These data were then represented as a graph where each node was a publication and each edge, the normalized co-citation frequency of a co-cited pair.  To cluster, we performed variable level clustering (VLC) followed by agglomerative clustering of the clusters resultant from VLC. To focus on major clusters, we also restricted agglomeration to clusters containing at least 10 nodes, which resulted in 666 clusters. Further agglomeration resulted in the development of a single large cluster (Fig~\ref{fig:merge_select}).  Consequently, the 20 largest clusters were used for further analysis and comparison with clusters generated by direct citation. The largest of these 20 clusters contained 1,276 nodes and the smallest contained 136, with total nodes summing to 8,042.

[Overall conclusions of cocitation based clustering]

@@

\begin{figure}[ht]
% Use the relevant command to insert your figure file.
% For example, with the graphicx package use
  \includegraphics[scale=0.45]{scopus_dblp_graclus3.pdf}
% figure caption is below the figure
\caption{Heat map for the clustering obtained by direct citation. The  y-axis (rows) correspond to topics, defined by Scopus characterizations, and the
x-axis (columns) represent the 20 different clusters.  
Each cluster is characterized by topics that label at least 10\% of the publications in the cluster.
 }
\label{fig:heatmap}       % Give a unique label
\end{figure}

@@

\begin{figure}[ht]
% Use the relevant command to insert your figure file.
% For example, with the graphicx package use
  \includegraphics[scale=0.45]{scopus_dblp_heatmap3.pdf}
% figure caption is below the figure
\caption{Heat map for the clustering obtained by direct citation. The  y-axis (rows) correspond to topics, defined by Scopus characterizations, and the
x-axis (columns) represent the 20 different clusters.  
Each cluster is characterized by topics that label at least 10\% of the publications in the cluster.
 }
\label{fig:heatmap}       % Give a unique label
\end{figure}
\newpage

>>>>

\title{Computer Science, Vignettes, and Citation Analysis}
%\thanks{Grants or other notes
%about the article that should go on the front page should be
%placed here. General acknowledgments should be placed at the end of the article.}
\subtitle{Salton and Bergmark Redux}
%\titlerunning{Short form of title}        % if too long for running head

@@

\begin{itemize}
\item understanding structure of communications within a field and with adjacent fields
\item extending Salton and Bergmark and staying high-level
\item combining article level and journal level 
\item two clustering methods
\item Graclus- all inclusive, unweighted edges
\item Co-citations, highly selective and weighted edges
\item intersection of Graclus and co-citation
\item close out
\end{itemize}

@@

Other classification systems are available such as the All Science Journal Classification (ASJC) developed and maintained by Scopus, the Web of Science research and categorical classifications from Clarivate Analytics, and the NSF classification system~\cite{nsf_classification} [add citations for Scopus and Web of Science]. Scopus and the Web of Science also include comprehensive bibliographies with citation links and proprietary unique identifiers for publications. All three systems rely on applying one or more journal-derived labels to articles.  A logical prediction, given the diversity of articles within journals, is limited specificity at the article level. Shu and colleagues recently noted in a comparative study of the Chinese Science Citation Database (CSCD) and the Web of Science, that 46\% of articles did not belong to the discipline of the journal they were published in~\cite{shu_comparing_2019}. Others have also discussed and critiqued disciplinary assignments using journal based classifications~\cite{wang_large-scale_2016,perianes-rodriguez_comparison_2017}. Article classification systems have been constructed, which escape some of the criticisms of journal-based classification~\cite{traag_louvain_2019,boyack_classification_2014,waltman_new_2012} but seem to be in the research domain rather than the domain of practical use [need to check on this].[May need another bridging sentence to the next paragraph].

@@

\subsection{Data}
 A stable release of the DBLP computer science bibliography~\cite{dblp_ref} consisting of 7,079,994 records was downloaded as dblp-2018-08-01.xml.gz. Just over 95\% of the publications within were published after 1996. Publications were parsed from the XML source file and loaded into a PostgreSQL database.  We have previously parsed the Scopus dataset of over 72,000,000 publications into a custom schema as part of implementing a larger data platform for research evaluation~\cite{GithubERNIE2019}. The total number of publications in Scopus that are labeled COMP with major subject area of Computer Science is 5,835,160. Records in the DBLP dataset were matched to Scopus identifiers using digital object identifiers (DOIs). This procedure resulted in a dataset of 2,685,356 DBLP publications with Scopus identifiers where 1,278,322 (47.6\%) were labeled as article and 1,407,034 (52.4\%) as conference proceedings.  References cited by these publications were then extracted from Scopus (7,129,006 records), resulting in a total of 8,000,411 nodes (publications and references) and 44,296,381 edges that represent citations within the dataset. This dataset is referred to as \emph{comp} (Fig~\ref{fig:ar_cp_annotation}, Table~\ref{tab:comp}).

@@

Figure~\ref{fig:heatmap} permits examination of these data from two perspectives: (i) rows: the clusters that map to a given ASJC minor subject area (ii) columns: ASJC minor subject areas that comprised at least 15\% of the publication in a cluster. Under these conditions of clustering and a threshold of 15\%, 31 of the 334 ASJC minor subject areas are represented. Unsurprisingly, the broad categories Computer Science Applications, Software, and Electrical and Electronic Engineering register in 16, 12, and 10 clusters respectively while Artificial Intelligence mapped to 7 different clusters. At the other end of the range, 12 of the 31 ASJC minor subject areas were each detected in a single cluster.

@@

To us, these data suggest that areas central to computer science in 2019 (Salton and Bergmark's historical triad of hardware, software, and theory) are more likely to be found in multiple clusters, whereas areas such as biology, chemistry, and statistics are more restricted in distribution.  A second inference is that, in some cases journal-based classification and our article clusters are similar in terms of article content (Hardware and Architecture). Thirdly, that the ASJC minor subject area, computer applications is relatively non-specific, and publications thus labeled are near ubiquitous (16/20).  Fourthly, that for the DBLP dataset, interactions with fields outside computer science are restricted to a single cluster or small numbers of clusters, e.g., cluster 17  and cluster 18.

We also explored mapping these 20 clusters assembled through direct citation to the CCS classification by manually matching the top 25-50 most heavily cited publications in each cluster to corresponding categories in the CCS. This was feasible in some cases, clusters 0 and 1 mapped reasonably well to the top level categories Hardware and Computer Systems Organization but much less so in the cases of other clusters. Further, the top cited papers often derived from biology and we were concerned that they were not representative of the cluster. The challenges inherent in labeling citation clusters at a high level of aggregation have been previously noted~\cite{waltman_new_2012}.

@@

\subsection{Clustering by Co-citation.} For an alternate examination of the data, we used co-citation frequencies to cluster the DBLP dataset. Unlike clustering by direct citation with Graclus, where every input publication is assigned to a cluster and every citation is weighted equally, clustering by co-citation is progressively selective, involves assigning weights to co-citations, and also accounts for weak inter-cluster interactions.

Co-citation was first described independently by Small and Marshakova in 1973~\cite{small_co-citation_1973,marshakova-shaikevich_co-citation_1973}, and provides insight into the emergence of new ideas derived from the association of previously independent ones.  Co-citation thus provides some degree of evolutionary insight and has been used for clustering publications~\cite{boyack_cocitation_2010,boyack_improving_2013,small_structure_1974,small_clustering_1985}.

As described in Methods, we first identified the top 10\% most-cited articles in the \emph{comp} dataset and then documented all co-cited pairs of their cited references that were also elements of \emph{comp}. The frequency of co-citation was then computed across all records in Scopus and normalized using Salton's correction. These co-cited pairs and the normalized co-citation frequencies were then represented as a graph where each node was a publication and each edge represented a co-citation, weighted by the normalized co-citation frequency.

For clustering, we applied variable level clustering (VLC) followed by 600 iterations of agglomeration of the clusters resultant from VLC since further rounds of agglomeration resulted in the development of a single large cluster (Fig~\ref{fig:merge_select}). To focus on major clusters, we also restricted agglomeration to clusters containing at least 10 nodes.  Consequently, the 20 largest clusters were used for further analysis and comparison with clusters generated by direct citation. The largest of these 20 clusters contained 1,276 nodes and the smallest contained 136, with total nodes summing to 8,042.

Fig~\ref{fig:heatmap_cocit} shows a heatmap in which clusters constructed by co-citation are mapped to Scopus ASJC minor subject areas. Mappings are only shown if they account for at least 15\% of the publications in a cluster. Unlike the earlier figure [insert figure reference] Unlike the case of clustering by direct citation, no cluster maps to more than three ASJC minor subject areas. We have to decide what to say about this figure but I think it suggests intra-cluster relationships for the most part.

@@

\begin{figure}[ht]
% Use the relevant command to insert your figure file.
% For example, with the graphicx package use
  \includegraphics[scale=0.45]{scopus_dblp_graclus3.pdf}
% figure caption is below the figure
\caption{Heat map for the clustering obtained by direct citation. The  y-axis (rows) correspond to topics, defined by Scopus characterizations, and the
x-axis (columns) represent the 20 different clusters.  
Each cluster is characterized by topics that label at least 15\% of the publications in the cluster.
 }
\label{fig:heatmap}       % Give a unique label
\end{figure}

@@

\begin{figure}[ht]
% Use the relevant command to insert your figure file.
% For example, with the graphicx package use
  \includegraphics[scale=0.45]{scopus_dblp_heatmap3.pdf}
% figure caption is below the figure
\caption{Heat map for the clustering obtained by direct citation. The  y-axis (rows) correspond to topics, defined by Scopus characterizations, and the
x-axis (columns) represent the 20 different clusters.  
Each cluster is characterized by topics that label at least 15\% of the publications in the cluster.
 }
\label{fig:heatmap_cocit}       % Give a unique label
\end{figure}
\newpage
